---
title: "Footnotes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Footnotes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = "#>"
)
```

```{css, echo = FALSE}
img {
  border: 0px;
}
```
*Capital* shouldn't be thought of as an isolated work of political economy, but
as part of an ongoing discourse as to the mechanics of capital and the solution
to a society based around capital's impersonal domination. It can be easy to
lose sight of that, reading *Capital* on its own 150 years later, without
knowing the context of the discussions at the time. Marx can get a little snarky
about the works of earlier political economists, and often uses the footnotes to
sarcastically and humourously point out the flaws in his peers' arguments. It's
fun to think of these conversations happening not in weighty tomes but over a
platform like twitter.

Here, I explore the footnotes of Capital: Which footnote is the longest (and how
many tweets would that thread be?)? Which footnote is the shortest? What
proportion of footnotes would fit in a tweet? Who would Marx tag in his tweets?

## Methods

```{r setup, echo=FALSE}
library(marxmywords)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

theme_marx <- function() {
  theme_classic() %+replace%
    theme(
      strip.background = element_rect(fill = "white", color = NA),
      axis.text = element_text(color = "black"),
      plot.caption = element_text(color = "grey50", hjust = 1, size = 6),
      plot.subtitle = element_text(size = 7.5, color = "grey30", hjust = 0)
    )
}

pretty_chapter_names <- function(v){
  fct <- paste("Chapter", 1:33)
  factor(paste("Chapter", v), levels = fct)
}
```

{marxmywords} provides the text for *Capital* as a data frame. For more on the
structure and source of this text, see `help(capital_vol1)`. For now, we are
only interested in the footnotes, so we will filter for those rows (each row
corresponds to one footnote). We will also do a little bit of tidying up of the
text: remove leading and trailing white space and the number associated with the
footnote. Finally, we will count the number of characters in each footnote.

```{r}
tidy_footnotes <- capital_vol1 %>%
  filter(section == "footnotes") %>%
  # remove leading white space
  mutate(text = gsub("^ *", "", text)) %>%
  # extract footnote number 
  mutate(
    ref = as.numeric(gsub("(*)\\. .*", "\\1", text)),
    text = gsub("^[0-9]+\\. *", "", text)
  ) %>% 
  # remove trailing white space
  mutate(text = gsub(" *$", "", text)) %>%
  mutate(
    chapter_names = pretty_chapter_names(chapter),
    n_characters = nchar(text)
  ) %>%
  # remove empty rows
  filter(n_characters >= 1 & !is.na(ref))
```

## Analysis

### The longest footnote in Capital
```{r fig.height = 4, fig.width = 6, echo=FALSE}

max_fn_per_chapter <- tidy_footnotes %>%
  group_by(chapter) %>%
  slice(which.max(n_characters)) %>%
  select(chapter_names, longest_fn = n_characters)

tidy_footnotes %>%
  left_join(max_fn_per_chapter, by = "chapter_names") %>%
  mutate(color_flag = n_characters == longest_fn) %>%
  ggplot() +
  aes(x = chapter_names) +
  geom_segment(
    aes(
      x = chapter_names, 
      xend = chapter_names,
      y = 0, yend = n_characters * color_flag
    ),
    color = "grey90",
    size = 2.2
  ) +
  geom_jitter(
    aes(color = color_flag,  y = n_characters),
    size = 0.1,
    width = 0.1
  ) +
  annotate(
    "point",
    x = forcats::fct_rev(max_fn_per_chapter$chapter_names), 
    y = max_fn_per_chapter$longest_fn,
    color = "red"
  ) +
  ylab("characters per footnote") +
  theme_marx() +
  theme(
    legend.position = "none",
    axis.ticks = element_blank(),
    panel.grid.major.y = element_line(color = "grey95"),
    axis.line.y = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_blank(),
    axis.text.x = element_text(size = 9, angle = 90, vjust = 0.5),
    plot.title = element_text(size = 12, hjust = 0)
  ) +
  scale_color_manual(values = c("black", "red")) +
  scale_size_manual(values = c(0.1, 1)) +
  scale_y_continuous(expand = c(0, 0, 0.1, 0.1)) +
  labs(
    title = "Footnotes in Capital: the long and the short",
    caption = "{marxmywords}"
  )

```
From the graph above, we can see the longest footnote is in Chapter 25, although
there are a few close competitors!

```{r }
longest_footnote <- tidy_footnotes %>%
  slice(which.max(n_characters)) %>%
  pull(text)

```

> `r longest_footnote`

So there you have it: Malthus provoked the longest diatribe.

### The shortest footnotes

The shortest footnote is unfortunately a little harder to identify.
Naively sorting for the fewest characters gets us a bunch of "l.c." ("last
cited") plus a reference to Petty (the economist, though it would be funny if
this was instead a commentary on the small mindedness of a statement).

```{r}
tidy_footnotes %>%
  arrange(n_characters) %>%
  head()

```


```{r fig.width=4,fig.height=3, echo=FALSE}
shortest_non_ref_fn <- data.frame(
  ref = c(144, 12, 30, 44, 2, 7),
  chapter = c(15, 7, 15, 25, 30, 3)
)

inner_join(tidy_footnotes, shortest_non_ref_fn, by = c("chapter", "ref")) %>%
  mutate(text = reorder(text, n_characters)) %>%
  ggplot() +
  aes( x = n_characters, y = text, color = text ) +
  geom_point() +
  geom_segment(aes(xend = 0, yend = text)) +
  geom_label(
    aes(x = 0, label = text),
    fill = "white",
    label.size = NA,
    label.padding = unit(0, "cm"),
    hjust = 0,
    nudge_x = 1,
    size = 3.2
  ) +
  scale_color_manual(values = c("red", rep("black", 5))) +
  scale_x_continuous(expand = c(0, 0, 0.1, 0.1)) +
  theme_marx() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_line(size = 0.4),
    axis.title.y = element_blank()
  ) +
  xlab("Number of characters") +
  labs(
    title = "Footnotes in Capital: shortest footnotes",
    caption = "{marxmywords}"
  )

```

### Tweetable Footnotes

From our analysis above, we know that the shortest non-citation footnote is 29
characters long. Twitter allows tweets up to 280 characters. How many fall in
between these limits?

```{r fig.height=4.5, fig.width=6, echo=FALSE}

twitter_text <- c(
  "just citations",
  "potentially great twitter material",
  "too long to tweet"
)

fn_fill_counts <- tidy_footnotes %>%
  mutate(fill_flag = case_when(
    n_characters <= 28   ~ twitter_text[1],
    n_characters > 280   ~ twitter_text[3],
    TRUE                 ~ twitter_text[2]
  )) %>%
  group_by(fill_flag) %>%
  tally() %>%
  mutate(fill_flag = factor(fill_flag, levels = twitter_text)) %>%
  arrange(fill_flag)

fn_fill_counts$xmid <- c(
  log10(28/2), 
  (log10(280)-log10(28))/2 + log10(28), 
  (log10(6852) - log10(280))/2 + log10(280)
)


p <- tidy_footnotes %>%
  ggplot() + 
  aes(x = n_characters) +
  geom_density() +
  scale_x_log10()

ggplot_build(p)$data[[1]] %>%
  mutate(fill_flag = case_when(
    x <= log10(28)   ~ twitter_text[1],
    x > log10(280)   ~ twitter_text[3],
    TRUE             ~ twitter_text[2]
  )) %>%
  mutate(fill_flag = factor(fill_flag, levels = twitter_text)) %>%
  ggplot() +
  aes(x = x, y = y, fill = fill_flag) +
  geom_area() +
  annotate(
    "segment",
    linetype = "dashed", size = 1, 
    x = log10(28), xend = log10(28),
    y = 0, yend = 0.45
  ) + 
  annotate(
    "text",
    x = log10(28),
    y = 0.5,
    size = 3.5,
    label = "everything < 29 characters\n is just a citation"
  ) +
  annotate(
    "segment",
    linetype = "dashed", size = 1, 
    x = log10(6852), xend = log10(6852),
    y = 0, yend = 0.14
  ) + 
  annotate(
    "text",
    x = log10(6852),
    y = 0.30,
    size = 3.5,
    label = "the longest\nfootnote\nwould be a\n25-tweet thread\nslamming\nMalthus"
  ) +
  annotate(
    "segment",
    linetype = "dashed", size = 1, 
    x = log10(280), xend = log10(280),
    y = 0, yend = 0.70
  ) + 
  annotate(
    "text",
    x = log10(280),
    y = 0.75,
    size = 3.5,
    label = "280 characters"
  ) +
  annotate(
    "text",
    label = paste0(fn_fill_counts$n, "\nfootnotes"),
    x = fn_fill_counts$xmid,
    y = 0.05,
    color = "black"
  ) +
  theme_marx() +
  theme(
    legend.position = c(0.15, 0.9),
    legend.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line.y = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title = element_blank()
  ) +
  scale_fill_manual(values = c("grey60", "#1DA1F2", "grey80")) +
  ylab("Quantity of footnotes") +
  xlab("Number of characters (log scale)") +
  labs(
    title = "Footnotes of Capital: what if Marx had twitter?",
    caption = "{marxmywords}"
  ) +
  scale_x_continuous(expand = c(0.01, 0.35)) +
  scale_y_continuous(expand = c(0, 0, 0.1, 0.1))

```

### Citation Content

Who would Marx tag in his tweets? Whose tweets would he be replying to? Adam
Smith and Ricardo spring to mind, perhaps first and foremost; who else makes it
into the top 10?

To answer this, I used spacyr, an R port of the python NLP package spacy, and
extracted all words tagged as proper nouns. The writers below all had at least
20 mentions across both the main text and footnotes. 

```{r fig.height=3,fig.width=4, echo=FALSE}
count_total_mentions <- function(word, text) {
  suppressWarnings(sum(stringr::str_count(text, word)))
}

philosophers <- c(
  "Smith",
  "Mill",
  "Marx",
  "Engels",
  "Ricardo",
  "Senior",
  "Malthus",
  "Ure",
  "Hunter",
  "Wakefield"
)

main_and_footnotes <- capital_vol1 %>%
  filter(section %in% c("footnotes", "body"))

philosopher_counts <- vapply(
  philosophers, 
  count_total_mentions, 
  main_and_footnotes$text,
  FUN.VALUE = numeric(1)
)


philosopher_counts %>%
  stack() %>%
  select(philosopher = ind, mentions = values) %>%
  mutate(philosopher = reorder(philosopher, mentions)) %>%
  ggplot() +
  aes(y = philosopher, x = mentions) +
  geom_point() +
  geom_segment(
    aes(xend = 0, yend = philosopher)
  ) +
  scale_x_continuous(
    expand = c(0, 0, 0.1, 0.1)
  ) +
  scale_color_manual(
    values = c(rep("black", 9), "red")
  ) +
  theme_marx() +
  theme(
    legend.position = "none",
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.y = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, size = 12)
  ) +
  xlab("# mentions") +
  labs(
    title = "Capital: cited writers by # mentions",
    caption = "{marxmywords}"
  )
  

```
(Note that I didn't distinguish between father and son James Mill and John
Stuart Mill. Some footnotes were added by Engels and signed with his full name,
and so would be counted in here, however others are signed F.E., and would not
be included in this count.)

Most of these aren't so surprising. Smith, Ricardo, the Mills, Malthus, and
Senior all wrote political economy works that Marx was directly critiquing
and/or building on. It also makes sense than Marx would cite his earlier works
(mostly *Zur Kritik der Politischen Oekonomie* and *Misère de la Philosophie*),
and those of his frequent co-author Engels.

A few heavily cited people stood out to me. Dr. Julian Hunter was commissioned
to investigate public health issues affecting the working class, and his
saddening findings are cited mainly in Chapter 25 (20/25 mentions). Politician
[Wakefield](https://en.wikipedia.org/wiki/Edward_Gibbon_Wakefield) ranks in the
top ten despite nearly all of his mentions (20/25) occurring in a single,
relatively short chapter. This chapter (chapter 33) discusses primitive
accumulation and colonialism, and Wakefield had a scheme of systematic
colonization named after him. [Dr Andrew
Ure](https://en.wikipedia.org/wiki/Andrew_Ure) wrote on the manufacturing
industry in England and is cited primarily in Chapter 15 (18/37).



