---
title: "Word Frequencies in Capital"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Word Frequencies in Capital}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = "#>"
)
```

```{css, echo = FALSE}
img {
  border: 0px;
}
```

What words come to mind when you think of *Capital*? Labour, surplus value, linens
and coats? Can we learn anything new from *Capital*, or distill its contents
better, by looking at word frequencies? In this analysis, I'll show some example
analyses of *Capital* using the [tidytext
framework](https://www.tidytextmining.com/index.html).

```{r setup, echo=TRUE}
library(marxmywords)
library(dplyr)
library(tidyr)
library(tidytext)
library(SnowballC)
library(ggplot2)
library(ggtext)
library(scales)
library(igraph)
library(ggraph)

theme_marx <- function() {
  theme_classic() %+replace%
    theme(
      strip.background = element_rect(fill = "white", color = NA),
      axis.text = element_text(color = "black"),
      plot.caption = element_text(color = "grey50", hjust = 1, size = 6),
      plot.subtitle = element_text(size = 7.5, color = "grey30", hjust = 0),
      axis.line.y = element_blank(),
      panel.background = element_rect(color = NA)
    )
}

pretty_chapter_names <- function(v){
  fct <- paste("Chapter", 1:33)
  factor(paste("Chapter", v), levels = fct)
}
```

## Pre-processing Methods

{marxmywords} provides the text for *Capital* in a data frame, with columns
indicating the section, part and chapter corresponding to the text in the "text"
column. The data frame has one row per paragraph line break, according to the
HTML markup at the [Marxists Internet
Archive](https://www.marxists.org/archive/marx/works/1867-c1/index.htm). For
more documentation about the structure and source of this data set, see
`help(capital_vol1)`.

For this analysis, I used only the body of the text, so first let's filter for
those lines of text. (I'll look at the footnotes in a later analysis.)

```{r}
capital_main <- capital_vol1 %>%
  filter(section == "body") %>% 
  # handle some annoying common cases not handled well by tokenization tools
  mutate(
    text = gsub("co-operat", "cooperat", text, ignore.case = T),
    text = gsub("C-M-C", "CMC", text, ignore.case = T),
    text = gsub("M-C-M", "MCM", text, ignore.case = T)
  )

head(capital_main) %>%
  tibble()
```

### Stop Words

Stop words are common words like "the" and "an" that aren't so insightful to
include in some analyses. Tidytext's `stop_words` data frame includes stop words
from three lexicons. I opted to use only the words in the "snowball" and "onix"
lexicon (468 words total). The SMART lexicon includes words like "value" that
are rather crucial for this particular text.

```{r stop_words}
data(stop_words)

stop_words <- stop_words %>%
  filter(lexicon %in% c("onix", "snowball"))

# Also add in a few common words that aren't so interesting:
custom_stop <- data.frame(
  word = c("day", "days", "time", "hour", "hours"),
  lexicon = "custom"
)

stop_words <- stop_words %>%
  bind_rows(custom_stop)
```

### Set up (tokenized) tidytext data frames

To look at word frequencies, we first have to split the the text into useful
individual units, or tokens, of text. For this, we will use
`tidytext::unnest_tokens()` to split the text into individual words and into
bigrams (neighbouring pairs of words).

I've also added a column for word stems; this will allow us to group
"commodities" and "commodity" and other stem-sharing words later on.

```{r create_base_dataframe}
capital_tokenized <- capital_main %>%
  unnest_tokens(word, text) %>%
  mutate(stem = wordStem(word))  %>%
  # remove numeric tokens (references, dates, etc)
  filter(!grepl("[0-9]+", word))

capital_unstopped <- capital_tokenized %>%
  anti_join(stop_words, by = "word")

capital_bigrams <- capital_main %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  filter(
    # remove stop words
    !word1 %in% stop_words$word & !word2 %in% stop_words$word,
    # remove numeric (they're kinda boring, but "12 hours" is pretty common)
    !grepl("[0-9]", word1) & !grepl("[0-9]", word2) 
  ) %>%
  drop_na()
```

## Analysis

### Chapters by word count

There's a lot of variability in chapter length! Chapter 15 is nearly 45,000
words long while Chapter 29 is just 541 words.

```{r wordcount_by_chapter, fig.height=4.5,fig.width=6}
# I'm including stop words in this wordcount since I think that gives a more
# informative "denominator", i.e. "frequency per N words" instead of "frequency
# per N non-stop words"

word_counts_per_chapter <- capital_tokenized %>%
  group_by(chapter) %>%
  summarize(wordcount = n())

word_counts_per_chapter %>%
  # annotate the longest chapters
  mutate(
    annotate_chapters = case_when(
      chapter == 1  ~ "Commodities",
      chapter == 3  ~ "Circulation of<br>Commodities",
      chapter == 10 ~ "The Working-Day",
      chapter == 15 ~ "Machinery and Modern Industry",
      chapter == 25 ~ "The General Law of<br>Capitalist Accumulation",
      TRUE          ~ NA_character_
    )
  ) %>%
  mutate(
    chapter_pretty = pretty_chapter_names(chapter),
    color_flag = is.na(annotate_chapters)
  ) %>%
  ggplot() +
  aes(x = chapter_pretty, y = wordcount, color = color_flag) +
  geom_point() +
  geom_segment(
    aes(xend = chapter_pretty, yend = 0)
  ) +
  ggtext::geom_richtext(
    aes(y = wordcount/2, label = annotate_chapters),
    fill = "white",
    label.size = NA,
    label.padding = unit(0, units = "cm"),
    size = 3.1,
    angle = 90
  ) +
  scale_color_manual(values = c( "red", "black")) +
  scale_y_continuous(
    expand = c(0, 0, 0.1, 0.1),
    label = comma
  ) +
  theme_marx() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.title.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_line(color = "grey90")
  ) +
  ylab("words per chapter") +
  labs(
    title = "Chapters of Capital by the numbers",
    caption = "{marxmywords}"
  )

```

### Most common words

Let's try answer the question posed at the top of the analysis: do the words that come to mind when you think of *Capital* actually rank as the most common words? Here, we will count word frequencies by word stem, which allows us to group together "labourer" and "labourers" and similar pairs. For display purposes, I'll show the most common word associated with each stem.

```{r wordcounts, fig.width = 4, fig.height =3}
subt <- "Word counts by stem (e.g., 'commodity' and 'commodities')"
counts_by_stem  <- capital_unstopped %>%
  # sum up counts across each word
  group_by(word, stem) %>%
  tally() %>%
  # sum up counts across words with the same stem
  ungroup() %>%
  group_by(stem) %>%
  mutate(stem_n = sum(n)) %>%
  # take the most common word for that stem for our example
  slice(which.max(n)) %>%
  ungroup() 

top_10_words <- counts_by_stem %>%
  arrange(desc(stem_n)) %>%
  slice(1:10) %>%
  mutate(word = reorder(word, stem_n))

top_10_words %>%
  ggplot() +
  aes(x = stem_n, y = word, colour = word) +
  geom_point() + 
  geom_segment(
    aes(xend = 0, yend = word)
  ) +
  scale_color_manual(
    values = c(rep("black", 9), "red")
  ) +
  theme_marx() +
  theme(
    legend.position = "none",
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  scale_x_continuous(
    expand = c(0, 0, 0.1, 0.1)
  ) +
  xlab("# occurrences") +
  labs(
    title = "Capital: top 10 words",
    subtitle = subt,
    caption = "{marxmywords}"
  )

```

The top ten words are... pretty much exactly what we might expect! Labour rises
to the top, its fair due given that it is the source of all capital.

The frequency of the words "value" and "form" are interesting. On their own,
they don't have too much meaning, but Marx uses them extensively in partnership
with other terms: surplus *value*, exchange *value*, money *form*, equivalent
*form*, *value* *form*. What can we learn by looking at word pairs?

### Most common word pairs

Let's look at which words occur most commonly in sequence.

```{r fig.width = 6, fig.height = 4.5}

top_n_bigrams <- capital_bigrams %>%
  group_by(word1, word2) %>%
  tally() %>%
  ungroup() %>%
  arrange(desc(n)) %>%
  slice(1:55) 

# take only combinations where words occur more than once
table_bigram <- table(c(top_n_bigrams$word1, top_n_bigrams$word2)) %>%
  stack() %>%
  select(word = ind, pairs = values)
just_once_words <- table_bigram$word[table_bigram$pairs == 1]

word_graph <- top_n_bigrams %>%
  filter(!(word1 %in% just_once_words & word2 %in%just_once_words)) %>%
  graph_from_data_frame()

p_layout <- ggraph(word_graph, layout = "fr")$data %>%
  left_join(table_bigram, by = c("name" = "word"))

p_layout %>%
  ggplot() +
  aes(x = x, y = y) +
  geom_edge_link(color = "grey80") +
  geom_node_point(
    aes(size = log(pairs, base = 10)),
    color = "red",
    alpha = 0.8
  ) +
  geom_node_text(
    aes(label = name, alpha = log(pairs, base = 6)),
    repel = TRUE
  ) +
  scale_alpha_continuous(range = c(0.6, 1)) +
  scale_size_continuous(range = c(2, 10)) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.caption = element_text(color = "grey50", hjust = 1, size = 7)
  ) +
  labs(
    title = "Concept Clusters in Capital",
    caption = "{marxmywords}"
  )
```

Which pairs of words occur most commonly in *Capital*? I would guess "surplus
value"!

```{r fig.width = 4, fig.height =3}
top_10_bigrams <- capital_bigrams %>%
  group_by(word1, word2) %>%
  tally() %>%
  ungroup() %>%
  arrange(desc(n)) %>%
  slice(1:10)

top_10_bigrams %>%  
  mutate(bigram = reorder(paste(word1, word2), n)) %>%
  ggplot() +
  aes(y = bigram, x = n, color = bigram) +
  geom_point() +
  geom_segment(
    aes(xend = 0, yend = bigram)
  ) +
  scale_x_continuous(
    expand = c(0, 0, 0.1, 0.1)
  ) +
  scale_color_manual(
    values = c(rep("black", 9), "red")
  ) +
  theme_marx() +
  theme(
    legend.position = "none",
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.y = element_blank()
  ) +
  xlab("# occurrences") +
  labs(
    title = "Capital: top 10 word pairs",
    caption = "{marxmywords}"
  )

```

What we see here is that "labour power" is actually the most common word pair,
surpassing "surplus value" by a surprising 36%! In fact, the word "labour"
occurs 2944 times – so 20% of the occurrences of the word "labour" are in the
context of "labour power".

Labour power is an abstraction of the quantity of (socially necessary) work a
labourer can perform as part of the social contract they enter into
with the owner of the means of production (a capitalist). Labourers sell not
their *labour* but their *labour power*. This difference is essential, since it
permits the extraction of surplus value. If labourers were paid by their labour
and not their labour power, there would be no surplus value.

> Whence, then, arises the enigmatical character of the product of labour, so
soon as it assumes the form of commodities? Clearly from this form itself. The
equality of all sorts of human labour is expressed objectively by their products
all being equally values; the measure of the expenditure of labour power by the
duration of that expenditure, takes the form of the quantity of value of the
products of labour; and finally the mutual relations of the producers, within
which the social character of their labour affirms itself, take the form of a
social relation between the products.

It's really neat that this crucial, but kind of tricky, concept of labour power
occurs more frequently than surplus value, a term more frequently encountered in
Marxist discussions that I've read.


