---
title: "Social Hell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Social Hell}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  fig.align = "center"
)
```

```{css, echo = FALSE}
img {
  border: 0px;
}
p.caption {
  font-size: 0.8em;
}
```


```{r setup, echo=FALSE}
library(marxmywords)
library(dplyr)
library(tidyr)
library(tidytext)
library(SnowballC)
library(ggplot2)
library(ggforce)
library(geomtextpath)

theme_marx <- function() {
  theme_classic() %+replace%
    theme(
      strip.background = element_rect(fill = "white", color = NA),
      axis.text = element_text(color = "black"),
      plot.caption = element_text(color = "grey50", hjust = 1, size = 6),
      plot.title = element_text(
        size = 10, 
        color = "#f20707", 
        face = "bold", 
        hjust = 0
      ),
      plot.subtitle = element_text(size = 8.5, color = "grey30", hjust = 0)
    )
}

pretty_chapter_names <- function(v){
  fct <- paste("Chapter", 1:33)
  factor(paste("Chapter", v), levels = fct)
}

custom_stop <- data.frame(
  word = c(
    # units
    "p.m", "a.m", "y", "lbs", "oz", "yards",
    # equation variables
    "b"
  )
)
```
> At the entrance to science, as at the entrance to hell, the demand must be
made: "Here you must abandon every suspicion; here must all your cowardice 
die."  
  - 1859 Preface, Marx, quoting Dante's Divine Comedy

> Capital is dead labour, that, vampire-like, only lives by sucking living
labour, and lives the more, the more labour it sucks.  
  - Capital Vol 1, Marx 1867

## Introduction

In *Capital*, Marx famously uses vivid metaphors, like monsters and vampires, to
describe the uniquely exploitative nature of capital. But, what if the entire
organization of the book was itself an infernal metaphor?

William Clare Roberts wrote a book, *Marxs Inferno*, that argues that Marx
structured his argument in *Capital* along the literary framework of Dante's
*Inferno* — yes, the 14th century poem about a guy traveling through Hell.
(*Marx's Inferno* argues a few other things, like that *Capital* should be read
as an answer to contemporary socialist debates, but this part about the Divine
Comedy is its ["weirdest and most ambitious
claim"](https://redsails.org/losurdo-and-roberts/) and the one I'll focus on
here.) 

In *Inferno*, Dante is guided by the ancient poet Virgil through the nine
circles of Hell. At the center of Hell, they encounter Satan devouring Judas and
other famous betrayers. They escape Hell by climbing down Satan's legs, and pass
through the center of the universe. In Roberts' reading of *Capital*, "Marx
(...) cast himself as a Virgil for the proletariat, guiding his readers through
the lower recesses of the capitalist economic order in order that they might
learn not only how this 'infernal machine' works, but also what traps to avoid
in their efforts to construct a new world."

Like comparing capital to a vampire draining the life force of labourers
communicates capital's unbounded parasitism, building *Capital* as an allusion
to Dante's *Inferno* helps communicate a few complex ideas. Marx's opponents
were divided between those preoccupied with the moralistic questions of
socialism who wished to avoid political economy and those who wished to stay
firmly within classical political economy. In this literary allusion, we
encounter moral failures (incontinence, violence, fraud, treachery), but Marx
instead rejects this moralistic framework, arguing that "it is capital, as a
system of all-around domination, that is responsible for these evils, not the
individuals dominated by capital." In comparing the proletariat with Dante, Marx
"emphasizes the necessity of going *through* political economy in order to get
*beyond* it", of confronting Satan/capital in order to escape their Hell
(Roberts, 2017).

I'll leave it to Roberts to defend how fond of Dante Marx was, how often Marx
structured his other writings around literary works, and how clues from the
drafting process suggest an intent to make his book *more* like *Inferno*. But
is there support for this theory, of a descent through Hell, in the choice of
language used?

Sentiment analysis is a type of natural language processing (NLP) method that
attempts to understand the emotions and other subjective information in text.
This kind of analysis is used, for example, by social media sites to flag
offensive content. If *Capital* is patterned as a descent through Hell, we might
expect the sentiment of the text to become increasingly negative. Let's test
that hypothesis!

## Methods

I removed stop words from the main text of *Capital* (e.g., "the", "a", and
other linker words) using the onix and snowball lexicons available through
`{tidytext}`. I tokenized the text into individual words, extracted the word
stem (i.e., made it so that words like "worship" and "worships" were treated
identically) and then ascribed one or more sentiments to a word-stem using the
[NRC lexicon](https://onlinelibrary.wiley.com/doi/10.1111/j.1467-8640.2012.00460.x).
This lexicon was developed through crowd-sourcing (via Mechanical Turk)
relationships between words and emotions, and has been widely used (>1800
citations on Google Scholar). I selected this lexicon because, of the four
lexicons available in the `{tidytext}` package, the NRC lexicon mapped to the
highest proportion of words. However, the same analysis conducted with the Bing,
AFINN and Loughran lexicons recapitulated the trends shown here.

For the positivity/negativity analysis, positive words were assigned a value of
1 and negative words were assigned a value of -1. Net sentiments were calculated
and normalized to the total number of words matched with a sentiment.  I
excluded words from consideration if they were considered both positive and
negative, or if no sentiment could be ascribed. For the assessment of other
sentiments, the sum of words across each sentiment was normalized to the word
count in each hell, and then normalized to the sentiment value of the first
circle of Hell (for clearer data visualization). I excluded the word "money"
from this analysis since it occurs with very high frequency; with "money"
included in the analysis, sentiments associated with money (i.e., the
emotions anger, anticipation, joy, surprise and trust) reflected primarily the
incidence of the word "money." 

```{r stop_words, echo=FALSE}
data(stop_words)

stop_words <- stop_words %>%
  filter(lexicon %in% c("onix", "snowball"))

# Also add in a few common words that aren't so interesting:
custom_stop <- data.frame(
  word = c(
    # units
    "day", "days", "time", "hour", "hours",
     "p.m", "a.m", "y", "lbs", "oz", "yards",
    # equation variables
    "b", "n"
  ),
  lexicon = "custom"
)

stop_words <- stop_words %>%
  bind_rows(custom_stop)

require_automated_dl <- !interactive() && inherits(
  try(textdata::lexicon_nrc(), silent = TRUE), "try-error"
)

if (require_automated_dl) {
  dir <- rappdirs::user_cache_dir("textdata")
  path <- file.path(dir, "nrc", "NRCWordEmotion.rds")
  textdata:::download_nrc(dirname(path))
  textdata:::process_nrc(dirname(path), path)
}

nrc <- textdata::lexicon_nrc()
nrc_pn <- nrc %>%
  filter(sentiment %in% c("positive", "negative")) %>% 
  group_by(word) %>%
  filter(n() == 1) %>%
  mutate(sentiment = ifelse(sentiment == "positive", 1, -1))
nrc_other <- nrc %>%
  filter(!sentiment %in% c("positive", "negative"))

```

```{r echo=FALSE}
capital_main <- capital_vol1 %>%
  filter(section == "body") %>% 
  # handle some annoying common cases not handled well by tokenization tools
  mutate(
    text = gsub("co-operat", "cooperat", text, ignore.case = T),
    text = gsub("C-M-C", "CMC", text, ignore.case = T),
    text = gsub("M-C-M", "MCM", text, ignore.case = T)
  ) 

capital_tokenized <- capital_main %>%
  unnest_tokens(word, text) %>%
  mutate(stem = wordStem(word))  %>%
  # remove numeric tokens (references, dates, etc)
  filter(!grepl("[0-9]+", word)) %>%
  anti_join(stop_words)

stem_word_map <- capital_tokenized %>%
  group_by(stem, word) %>%
  tally() %>%
  ungroup() %>%
  group_by(stem) %>%
  slice(which.max(n)) %>%
  select(-n)


```

```{r}
marxs_inferno <- data.frame(
  part = 1:8,
  hell = c("Styx", rep("Dis", 2), rep("Malebolge", 4), "Cocytus")
) %>%
  mutate(
    hell = factor(hell, c("Styx", "Dis", "Malebolge", "Cocytus"))
  )

marxs_inferno_tokenized <- capital_tokenized %>%
  left_join(marxs_inferno, by = "part")

```

## The Circles of Hell In Capital

```{r fig.height = 5, fig.width = 5, fig.cap="Figure 1: Schematic illustrating similarities between Dante's Inferno and Capital.\n(Admittedly I just wanted an excuse to use geomtextpath)"}

greys <- c("grey70", "grey54", "grey38", "grey22")


# make black and white, with a white "out"
fig1 <- data.frame(
  hell = c("Styx", "Dis", "Malebolge", "Cocytus"),
  sins = c("Incontinence", "Violence", "Fraud", "Treachery"),
  circles = c("circles 1-5", "circles 6-7", "8th circle", "9th circle"),
  chapters = c("1-3", "4-11", "12-25", "26-33"),
  subject = c(
    "Commodities, exchange, money",
    "Capital, capitalist exploitation",
    "Capitalist mode of production & accumulation",
    "Primitive accumulation of capital"
  )
)

create_arcs <- function(r, hell) {
  arcs <- seq(r, 0, length.out = 1000)
  data.frame(
    x = -arcs,
    y = sqrt(r**2 - arcs **2),
    text = hell
  )
}

hell_arc_labels <- bind_rows(
  Map(create_arcs, 5:2 - 0.15, paste0(fig1$hell, ": ", fig1$sins)),
  create_arcs(1 + 0.15, "Grapple with Satan to..."),
  create_arcs(1 - 0.15, "...escape Hell"),
  create_arcs(1 - 0.35, "(socialism)")
)

capital_arc_labels <- bind_rows(
  Map(create_arcs, 5:2 - 0.35, paste0(fig1$subject)),
  Map(
    create_arcs, 5:2 - 0.55, 
    paste0(fig1$circles, ", chapters ", fig1$chapters)
  )
)

radii <- c(1, 1, 1, rep(1/2, 2), rep(1/5, 5))

radiuses <- data.frame(
  r = cumsum(radii), 
  r0 = c(0, cumsum(radii)[-10])
) 

radiuses %>% 
  ggplot() +
  geom_arc_bar(
    aes(
      r = r, r0 = r0, x0 = 0, y0 = 0, 
      start = -pi/2, end = 0, 
      fill = factor(r),
      color = factor(r)
    ),
    linewidth = 1,
    alpha = 0.85
  ) +
  geom_textpath(
    data = hell_arc_labels, 
    aes(x = x, y = y, label = text),
    linewidth = 0,
    size = 3,
    vjust = 0.5,
    face = "bold"
  ) +
  geom_textpath(
    data = capital_arc_labels, 
    aes(x = x, y = y, label = text),
    linewidth = 0,
    size = 2.8,
    vjust = 0.5
  ) +
  annotate(
    "text",
    x = -5, y = 5,
    hjust = 0,
    label = "Marx's Inferno",
    color =  "#f20707", face = "bold"
  ) +
  annotate(
    "text",
    x = -5, y = 4.8,
    hjust = 0, size = 3.2,
    label = "Parallels between Dante's Inferno & Capital",
    color =  "grey30"
  ) +
  annotate(
    "text",
    x = -5, y = 4.6,
    hjust = 0, size = 2.2,
    label = "Adapted from Roberts (2017)",
    color =  "grey30"
  ) +
  theme_void() +
  coord_fixed() +
  scale_fill_manual(
    values = rev(c(rep(greys[1], 5), rep(greys[2], 2), greys[3], greys[4], "white"))
  ) +
  scale_color_manual(
    values = rev(c(rep(greys[1], 5),  rep(greys[2], 2), greys[3], greys[4], "white"))
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 10, color = "#f20707", face = "bold"),
    plot.subtitle = element_text(size = 9, color = "grey30"),
    plot.caption = element_text(color = "grey50", hjust = 1, size = 6)
  ) 

```

Before we get into the results, let's quickly recap Roberts' argument for the
similarities between *Capital* and *Inferno*. Roberts breaks both texts into
four sections, as summarized in the figure above.

1. **Styx** (circles 1-5, chapters 1-3):
*Sins of incontinence; commodity fetishism*.  
The impersonal domination of market-based societies allows for free movement of
capital and products rather than control of capital/products by its producers.
2. **Dis** (circles 6-7, chapters 4-11): 
*Sins of violence; capitalist exploitation.*  
   The relationship between capitalists and their workers is inherently one of
*force*, as the boss controls and uses the workers to extract as much surplus
value as possible from the labour power sold by the labourer.
3. **Malebolge** (8th circle of hell, chapters 12-25): 
*Sins of fraud; the capitalist mode of production.*  
   While capitalism may promise improved productivity in the form of 
mechanization, co-operation, division of labour and other efficiencies of scale,
it delivers little of these gains to the workers but instead produces alienation
and miserable working conditions.
4. **Cocytus** (9th circle of hell, chapters 26-33): 
*Sins of treachery; primitive accumulation*.  
   Betrayals presented here include: (a) the expropriation of land from the
peasants in violation of the mutual obligation between serfs and their lords,
(b) the betrayal of the aristocrats in their new-found wealth by capital as the
capitalists/bourgeoisie are freed by the abolition of feudalism, (c) the
treachery of the state, acting as the agent of capital to do its "dirty" work,
and (d) that workers attempting to wield capitalism for their own benefit will
inevitably find themselves betrayed by the nature of capital.

It is only by moving *past* capitalism, by challenging Satan, that labourers can
free themselves from oppression:

> If Dante must confront Satan in order to escape his realm, then so must the
labouring classes confront capital in order to escape the social Hell. Instead
of trying to create their own capital, the labourers must realize that capital
is a wealth that betrays and turns against its creators. (Roberts, 2017)

## Positive and Negative Sentiments in Capital

Leaving the positive and negative aspects of capitalism to Marx to explain, what
words in *Capital* most contribute to positive or negative sentiment?

```{r fig.height=3,fig.width=4,fig.cap="Figure 2: Fall contributes the most to negative sentiments, although it's my favourite season." }
capital_tokenized %>%
  left_join(nrc_pn, by = "word") %>%
  filter(!is.na(sentiment)) %>%
  group_by(stem, sentiment) %>%
  tally() %>%
  left_join(stem_word_map) %>%
  ungroup() %>%
  mutate(
    contribution = sentiment * n,
    binary = ifelse(
      sentiment > 0, 
      "most positive", 
      "most negative"
    )
  ) %>%
  group_by(binary) %>%
  arrange(desc(abs(contribution))) %>%
  slice(1:6) %>%
  mutate(pos = row_number()) %>%
  mutate(word = reorder(word, abs(contribution))) %>%
  ggplot() +
  aes(x = contribution, y = word) +
  facet_wrap(~binary, scales = "free") +
  geom_point() +
  geom_segment(
    aes(yend = word, xend = 0)
  ) +
  geom_label(
    #aes(x = contribution/2, label = word),
    aes(x = 0, label = word,  hjust = ifelse(contribution > 0, 0, 1)),
    fill = NA,
    vjust = -0.1,
    label.size = NA,
    label.padding = unit(0.25, "lines"),
    size = 3.5
  ) +
  theme_marx() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title = element_blank(),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line("grey50"),
    axis.text.x = element_text(size = 8, color = "grey50"),
    strip.text = element_text(size = 12, face = "bold"),
  ) +
  labs(
    title = "Sentiments in Capital",
    subtitle = "Which words contributed the most to sentiment analysis?"
  )
  

```

From this graph, a couple limitations of this approach are clear: 

1.  *Sentiments are ascribed to individual words.* Words like "increase" and
"fall" are associated with positive and negative sentiment, respectively, but
should be considered in context. The phrase "an increase in exploitation" would 
have a net neutral sentiment (plus one for increase, minus one for
exploitation), whereas it doesn't take a particularly discerning reader to
interpret that phrase as negative. I repeated the analysis with words like
"increase", "decrease", "fall" and "rise" excluded, and trends remained the
same. More sophisticated algorithms than what I used here could address this
shortcoming, but this type of classification method is fairly common.

2.  *Sentiments are ascribed using popular interpretations of the word*. For
example, words like "money" are associated with positive sentiments. This isn't
necessarily *wrong*, having money is better than having no money. But one of the
key conclusions in *Capital* is that money inevitably arises from the
circulation of commodities (a concept with negative sentiments in *Capital*).
When Marx writes "circulation sweats money from every pore", money isn't being
used positively; it's a gross by-product. If we were ascribing sentiment to the
word "money" in the way it is used in *Capital*, a neutral or negative sentiment
might be closer to the "intended" sentiment of the text.

Perhaps the clearest example of differences between popular sentiment and
Marxist sentiment is the word "capitalist", the third highest contributor to
positive sentiment in *Capital*. Let's look at sentiments associated with
similar words...

```{r }
nrc_pn %>%
  filter(word %in% c("socialist", "communist", "capitalist")) %>%
  knitr::kable()
```

Like with any crowd-sourced data sets, existing social biases are propagated in
sentiment lexicons like the one used here. (A finding probably all too familiar
to those who've run into issues with automated content moderation on social
media sites.)

Let's keep these limitations in mind as we test our original hypothesis: do the
sentiments in *Capital* match a journey through Hell?

## Change in text positivity during the descent through hell

In Styx, we encounter sins of incontinence, and Marx presents value theory,
exchange and money. This hell is associated with the most positive sentiments.
From there, we decline in positivity into Dis and Malebolge. These circles of
hell are marked in particular by the very negative chapters 10, 15 and 25 — the
"historical" chapters that describe in detail the terrible living and working
conditions of British workers in the 19th century and the legal battles for
labour rights. We then encounter a massive drop in positivity as we descend into
the ninth circle of hell and study the treachery of primitive accumulation. This
relationship is statistically significant, for those who care about those sorts
of things (p = 0.0002).

```{r fig.height=4.5, fig.width=6, fig.cap="Figure 3: Chapter 10 made me feel absolutely miserable. It's nice when data validates my emotions."}
capital_sentiment <- marxs_inferno_tokenized %>%
  left_join(nrc_pn, by = "word") %>%
  filter(!is.na(sentiment)) 

sentiments_by_chapter <- capital_sentiment %>%
  group_by(hell, chapter) %>%
  summarize(
    sentiment = sum(sentiment)/n()
  ) %>%
  ungroup() %>%
  group_by(hell) %>%
  mutate(
    med = median(sentiment),
    mn = mean(sentiment)
  ) %>%
  ungroup() %>%
  mutate(
    chap_lab = ifelse(abs(sentiment - mn) > 0.15 , paste("Chapter", chapter), ""),
    chap_y   = 0.5 * (sentiment + mn)
  )

sentiments_overall <- capital_sentiment %>%
  summarize(
    sentiment = sum(sentiment)/n()
  )

hell_segments <- sentiments_by_chapter %>%
  group_by(hell) %>%
  summarize(
    xstart = min(chapter) - 0.3,
    xend   = max(chapter) + 0.3,
    xmid   = mean(chapter)
)


sentiments_by_chapter %>%
  left_join(hell_segments, by = "hell") %>%
  ggplot() +
  aes(x = chapter, y = sentiment, color = hell) +
  geom_hline(yintercept = sentiments_overall$sentiment, color = "grey90") +
  geom_segment(
    aes(
      xend = chapter,
      yend = mn
    )
  ) +
  geom_point() +
  geom_segment(
    aes(
      x    = xstart,
      xend = xend,
      y    = mn,
      yend = mn
    ),
    size = 1,
    alpha = 0.6
  ) +
  geom_text(
    aes(label = chap_lab, y = chap_y),
    angle = 90,
    size = 3,
    vjust = -0.4
  ) +
  annotate(
    "text",
    x = hell_segments$xmid, y = 0.7, label = hell_segments$hell,
    color = c("#f20707", "#bf0d0d", "#540505", "#330505")
  ) +
  annotate(
    "text",
    x = 0,
    y = c(0.1, -0.1) + sentiments_overall$sentiment,
    label = c("more positive", "more negative"),
    angle = 90, hjust = 0.5, vjust = -1,
    color = "grey40"
  ) +
  theme_marx() +
  scale_color_manual(
    values = c("#f20707", "#bf0d0d", "#540505", "#330505")
  ) +
  theme(
    legend.position = "none",
    axis.line = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank()
  ) +
  labs(
    title = "Marx's Inferno: sentiments in Capital",
    subtitle = "Sentiments expressed become more negative descending through the circles of hell"
  ) +
  scale_y_continuous()



```
```{r linear_model}
by_chapter <- lm(sentiment ~ chapter, sentiments_by_chapter)
```

## Change in sentiments by hell

In addition to positive/negative, the NRC sentiment lexicon also associates
words with one or more of eight emotions: anger, anticipation, disgust, fear,
joy, sadness, surprise, trust. In the *Marx's Inferno* framework, we might
expect increasing anger, disgust and fear as we descend through Hell. Based on
the unique characteristics of each hell, we might also expect higher surprise
and a loss of trust during the last two circles, which deal with fraud and
betrayal.

```{r fig.height = 3, fig.width = 4, fig.cap="Figure 4: Change in sentiment across Hell for all 8 sentiments in the NRC data set (excluding positive/negative)."}

marxs_inferno_tokenized %>%
  left_join(nrc_other, by = "word") %>%
  group_by(hell) %>%
  mutate(word_count = n()) %>%
  ungroup() %>%
  filter(
    !is.na(sentiment),
    !word %in% c("money")
  ) %>%
  group_by(hell, sentiment, word_count) %>%
  tally() %>%
  ungroup() %>%
  #normalize counts per hell to words per hell
  mutate(freq = n/word_count) %>%
  # normalize frequency to Styx (for nicer plotting)
  group_by(sentiment) %>%
  mutate(styx_freq = ifelse(hell == "Styx", freq, NA)) %>%
  fill(styx_freq, .direction = "downup") %>%
  mutate(freqn = freq/styx_freq) %>%
  ggplot() +
  aes(x = hell, y = freqn, group = 1) +
  geom_line() +
  facet_wrap(~sentiment, nrow = 2) +
  theme_void() +
  scale_y_continuous(
    limits = c(0, NA),
    breaks = c(0, 1, 2, 3, 4)
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 9),
    axis.text.y = element_text(size = 8, color = "grey30"),
    axis.title = element_blank(),
    panel.spacing.y = unit(1, "lines"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(size = 10, color = "#f20707", face = "bold"),
    plot.subtitle = element_text(size = 9, color = "grey30"),
    plot.caption = element_text(color = "grey50", hjust = 1, size = 6),
    strip.text = element_text(size = 10, vjust = 1)
  ) +
  labs(
    title = "Marx's Inferno",
    subtitle = "Sentiment in Capital by circle of Hell\n"
  )
  
```
Breaking down the words in *Capital* into these eight emotional categories, we
see a very strong increase in the negative emotions anger, disgust, fear and
sadness (275%, 170%, 150%, 152% increase in the final circle of Hell relative to
the initial circles of Hell). The uptick in joy is intriguing, but the effect
size is not nearly as large as for the other emotions (65% higher in Cocytus
relative to Styx). Overall, these trends are consistent with what we saw above
in terms of declining positivity over the course of *Capital*.

The emotions surprise and trust remain flat throughout *Capital*, counter to my
hypothesis that chapters associated with fraud and betrayal might be associated
with shock and loss of faith. While this finding doesn't necessarily provide
additional support to the arguments presented in *Marx's Inferno*, perhaps it
was too much to demand this level of emotional granularity in the text. After
all, capital's betrayal of workers was not a particularly radical or unexpected
claim to make even in 1867.

## Conclusions

Over the course of *Capital*, the sentiments expressed in the text become
increasingly negative, consistent with Dante's journey through the circles of
Hell. Similarly, anger, disgust, fear and sadness monotonically increased in
intensity over the course of the book. While the approach here is fairly simple
and is based on popular sentiments associated with words rather than
context-specific sentiments, these findings lend support to the arguments
presented in *Marx's Inferno*.

> Like Dante's Virgil, Marx tells his charges to disregard social Hell's command
— "Abandon ever hope, you who enter"  — ordering them instead to "abandon every
suspicion." And, like the pilgrim in Dante's poem, the socialists who accept
Marx's guidance are supposed to be so transformed by their journey that they
will be able to withstand the purgatory through which the revolution was
currently (and recurrently) travelling.  
 - Roberts, 2017

> [My Guide] first, I second, still ascending held  
  Our way until the fair celestial train  
  Was through an opening round to me revealed:  
  And, issuing thence, we saw the stars again.  
  - Dante's Inferno, 14th century
